---
title: "KSB620 SA2 – Visual Summaries"
author: "Tazvitya Aubrey Chihota"
format: pdf
---

## Introduction
 
The goal of this report is to apply effective data presentation techniques by producing seven high-quality visuals and descriptive tables using the **Real or Fake Job Postings** dataset.  

The dataset contains job listings with various attributes such as employment type, salary range, required experience, telecommuting availability, and a binary indicator showing whether a posting is **fraudulent** or **legitimate**.  

The analysis explores these variables through:
- Two **univariate plots** to describe single-variable distributions.  
- One **bivariate**, one **trivariate**, and one **quadrivariate** plot to examine relationships between multiple variables.  
- Two **descriptive tables** (one categorical and one numeric) to summarize key data characteristics.  

All visuals are generated using **R and the tidyverse**, following the *Effectively Presenting Data* principles — emphasizing clarity, simplicity, and accessibility. Each visualization is accompanied by a brief interpretation highlighting meaningful insights about employment trends and fraudulent activity patterns.

```{r setup}
# Load libraries
library(tidyverse)
library(janitor)
library(scales)
library(knitr)

# Load dataset
jobs <- read_csv("real_fake_job_postings.csv") %>% clean_names()

# Prepare salary variables
jobs <- jobs %>%
  separate(salary_range, into = c("min_salary", "max_salary"), sep = "-", remove = FALSE, fill = "right") %>%
  mutate(across(c(min_salary, max_salary), ~ as.numeric(str_remove_all(., "[^0-9]")))) %>%
  mutate(avg_salary = (min_salary + max_salary) / 2) %>%
  mutate(across(c(employment_type, required_experience, industry, fraudulent),~ replace_na(., "Unknown")))

```

```{r}
#| label: Univariate Plot (Categorical Variable)
ggplot(jobs, aes(x = employment_type, fill = employment_type)) +
  geom_bar() +
  theme_minimal(base_size = 12) +
  labs(title = "Distribution of Employment Types",
       x = "Employment Type", y = "Number of Postings") +
  theme(legend.position = "none")

```
## Most job postings are full-time positions, showing a dominance of stable employment over part-time or contract work.

```{r}
#| label: Univariate Plot (Numeric Variable)
# Filter valid salaries to remove NAs and zeros
salary_data <- jobs %>%
  filter(!is.na(avg_salary), avg_salary > 0, avg_salary < 500000)

# Histogram of average salaries
ggplot(salary_data, aes(x = avg_salary)) +
  geom_histogram(binwidth = 10000, fill = "#0072B2", color = "white", alpha = 0.8) +
  scale_x_continuous(labels = comma) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Distribution of Average Salaries in Job Postings",
    x = "Average Salary (USD)",
    y = "Number of Job Postings"
  ) +
  geom_vline(aes(xintercept = median(avg_salary, na.rm = TRUE)), color = "red", linetype = "dashed", size = 1)


```
## The distribution is right-skewed, with most jobs offering salaries below $70,000 and a few highly paid outliers.

```{r}
#| label: Bivariate Plot (Categorical + Numeric)
# Use cleaned salary data for valid salary values
ggplot(salary_data, aes(x = employment_type, y = avg_salary, fill = employment_type)) +
  geom_boxplot(outlier.alpha = 0.3, color = "gray20") +
  scale_y_continuous(labels = comma, limits = c(0, 200000)) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Average Salary by Employment Type",
    x = "Employment Type",
    y = "Average Salary (USD)"
  ) +
  theme(legend.position = "none") +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "yellow")


```

## Full-time positions show the widest salary spread and the highest average pay, while internships and temporary roles remain at the lower end.
```{r}
#| label: Trivariate Plot (Two Numeric + One Categorical)
ggplot(jobs, aes(x = telecommuting, y = has_company_logo, color = as.factor(fraudulent))) +
  geom_jitter(alpha = 0.4) +
  scale_color_manual(values = c("0" = "#2c7fb8", "1" = "#de2d26"),
                     labels = c("Legit", "Fraudulent")) +
  theme_minimal(base_size = 12) +
  labs(title = "Telecommuting and Company Logo vs Fraudulent Postings",
       x = "Telecommuting (0=No, 1=Yes)",
       y = "Has Company Logo (0=No, 1=Yes)",
       color = "Posting Type")

```

## Fraudulent postings are more common among listings that lack company logos and allow telecommuting.

```{r}
#| label: Quadrivariate Plot (4 Variables)
# Clean and simplify experience labels
jobs_cleaned <- jobs %>%
  filter(required_experience != "Unknown", employment_type != "Unknown") %>%
  mutate(
    required_experience = str_to_title(required_experience),
    required_experience = recode(required_experience,
      "Internship" = "Intern",
      "Entry Level" = "Entry",
      "Associate" = "Assoc",
      "Mid-Senior Level" = "Mid-Senior",
      "Director" = "Dir",
      "Executive" = "Exec"
    ),
    required_experience = fct_infreq(required_experience)
  )

# Improved quadrivariate plot
ggplot(jobs_cleaned, aes(x = required_experience, fill = as.factor(fraudulent))) +
  geom_bar(position = "fill") +
  facet_wrap(~ employment_type, ncol = 3) +
  scale_fill_manual(values = c("0" = "#4daf4a", "1" = "#e41a1c"),
                    labels = c("Legit", "Fraudulent")) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Fraud Rate by Experience Level and Employment Type",
    x = "Required Experience Level",
    y = "Proportion",
    fill = "Posting Type"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    strip.text = element_text(face = "bold", size = 11),
    legend.position = "bottom"
  )

```
## Fraudulent jobs concentrate in entry-level full-time listings, suggesting scammers target inexperienced applicants.
```{r}
#| label: Descriptive Table – Categorical Variable
cat_table <- jobs %>%
  count(employment_type) %>%
  mutate(Percent = percent(n / sum(n)))
kable(cat_table, caption = "Table 1: Distribution of Employment Types")

```
## Over half of postings are full-time roles, confirming a high demand for regular employment contracts.

```{r}
#| label: Descriptive Table – Numeric Variable
num_table <- jobs %>%
  summarize(
    Mean_Salary = mean(avg_salary, na.rm = TRUE),
    Median_Salary = median(avg_salary, na.rm = TRUE),
    SD_Salary = sd(avg_salary, na.rm = TRUE),
    Min_Salary = min(avg_salary, na.rm = TRUE),
    Max_Salary = max(avg_salary, na.rm = TRUE)
  )
kable(num_table, caption = "Table 2: Summary Statistics for Average Salary")

```
## The salary range shows high variance, indicating that the dataset includes both entry-level and high-paying roles.
